---
title: "Exploring data with R tutorial"
author: "Erin W Dickie"
date: "September 11, 2016"
output: html_document
---

[Return to SCWG Home](https://camh-scwg.github.io/compucool-oct-2016/)

-----

Please download the following files:


1. [messy_demographic.csv](data/messy_demographic.csv) 
2. [messy_cognitive.csv](data/messy_cognitive.csv)
3. [messy_genotype.csv](data/messy_genotype.csv)

----

We need 'ggplot2', 'dplyr', 'tidyr', and 'broom' packages
If you haven't already installed these:

Load the packages

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(broom)
library(knitr)
```


The data should be merged and ready to go from day 1. If not, here's the code for it:

```{r eval = FALSE}
data1 <- read.csv("~/Downloads/messy_demographic.csv")  # put in the location of the downloaded data file 1
data2 <- read.csv("~/Downloads/messy_cognitive.csv")  # put in the location of the downloaded data file 2
data3 <- read.csv("~/Downloads/messy_genotype.csv")
```
```{r echo = FALSE}
data1 <- read.csv("data/messy_demographic.csv")  # put in the location of the downloaded data file 1
data2 <- read.csv("data/messy_cognitive.csv")  # put in the location of the downloaded data file 2
data3 <- read.csv("data/messy_genotype.csv")
```


```{r}
data1[data1==""] <- NA
data1[data1=="missing"] <- NA
data1[data1=="9999"] <- NA

data1$age <- as.numeric(as.character(data1$age))
data1$ethnicity <- factor(data1$ethnicity,levels=c("Cauc","AA","As","In","Other"))
data1$sex <- factor(data1$sex, levels=c(0,1), labels=c("Male","Female"))
data1$dx <- factor(data1$dx, levels=c(0,1), labels=c("Control","Case"))

data2[data2==""] <- NA
data2[data2=="missing"] <- NA
data2[data2=="9999"] <- NA
data2$cog1 <- as.numeric(as.character(data2$cog1))
data2$cog2 <- as.numeric(as.character(data2$cog2))
data2$cog3 <- as.numeric(as.character(data2$cog3))
data2$subID <- gsub(data2$subID,pattern="subject",replacement="SUB_")

data3[data3==""] <- NA
data3[data3=="missing"] <- NA
data3[data3=="9999"] <- NA
data3$genotype <- factor(data3$genotype, levels=c(0,1,2), labels=c("AA","AG","GG"))
data3$subID <- gsub(data3$subID,pattern="subject",replacement="SUB_")

alldata <- merge(data1,data2,by.x="subject_ID",by.y="subID")
alldata <- merge(alldata,data3,by.x="subject_ID",by.y="subID")

## Write merged dataframe as new .csv
write.csv(alldata, "alldata.csv", row.names=F)
```
## RESEARCH QUESTION 3: 
#### cog1 ~ dx (t-test)



```{r}
t.test(data=alldata, cog1 ~ dx)
```

Wow we have a result!!
But one of my favorite things output R is that the statistical output can be saved as an object!!
```{r}
my.t.result <- t.test(data=alldata, cog1 ~ dx) # saves to output to my.t.result

print(my.t.result)       ## prints to output to the console
my.t.result$statistic    ## gets us the t statistic!
my.t.result$parameter    ## the degrees of freedom
my.t.result$p.value      ## gets us the p-value

round(my.t.result$statistic,2) ## we can these numbers using the "round" function
```

Let's put these three together into something we might want to report in our paper

```{r}
my.t.results.txt = paste0('t(',
                        round(my.t.result$parameter,1),
                        ') = ',
                        round(my.t.result$statistic,2), ', p = ',
                        round(my.t.result$p.value, 7))

my.t.results.txt
```


View this as basic boxplots (two ways: base package and ggplot2)

```{r}
ggplot(data=alldata, aes(y=cog1,x=dx)) + 
	geom_boxplot()
```

#### Let's make it fancier:

First - let's deal with the NA's we don't want to plot - let's remove them from the plotting dataset

```{r}
data.toplot <-filter(alldata, !is.na(cog1), !is.na(dx))

ggplot(data.toplot, aes(y=cog1,x=dx)) + 
	geom_boxplot(outlier.shape=NA) + ## removes outliers from boxplot layer
	geom_jitter(alpha=0.5)           ## add dots on top

## even fancier - let's add a title, and annotation and label the axes
ggplot(data.toplot, aes(y=cog1,x=dx)) + 
  geom_boxplot(outlier.shape=NA) + 
  geom_jitter(alpha=0.5) +
  labs(title="Effect of Diagnosis on Cog Score #1",
       y = "Cognitive Test 1",
       x = "Diagnosis") +
  annotate("text", label = my.t.results.txt, x = 1, y = 21) +
  theme_bw()
```

#### NOW LET's save our plot!!!

**Note**: we can start by using the "Export" button in the plots tab..
```{r eval=FALSE}
ggsave('figure1_ttestresults.pdf', width = 5, height = 5)
```

## Let's make a diagnosis by cognition table
```{r}
my.stats.table <- summarise(alldata,
                            "Mean" = mean(cog1, na.rm = T),
                            "St Dev" = sd(cog1, na.rm = T))

my.stats.table <- alldata %>% 
                  group_by(dx,sex) %>%
                  summarise("Mean" = mean(cog1, na.rm = T),
                            "St Dev" = sd(cog1, na.rm = T))

kable(my.stats.table)
```

## reshaping data to look at all three cognitive scores 

### but let's look at all three subscales....
```{r}
## let's add a rick carrier
alldata$riskcarrier[alldata$genotype=="AG" | alldata$genotype=="GG"] <- "carrier"
alldata$riskcarrier[alldata$genotype=="AA"] <- "non-carrier"
alldata$riskcarrier <- as.factor(alldata$riskcarrier)

alldata_melted <- gather(alldata, cog_var, cognitive.score, 
                  -subject_ID, -age, -sex, -ethnicity, -dx, -genotype, -riskcarrier)

toplot <- filter(alldata_melted, !is.na(cognitive.score), 
                                 !is.na(dx), 
                                 !is.na(riskcarrier))

ggplot(toplot, aes(y=cognitive.score,x=age,color=riskcarrier)) + 
  geom_point() + 
  geom_smooth(method=lm) + 
  facet_wrap(~cog_var, scales = "free")
```

## LAST CHALLENGE: can you use dplyr to present these statistics in a table??

LAST BUT NOT LEAST, running a loop to screen for significance at multiple tests (manhattan plot)

```{r}
## add stargazer descriptive stats table?
cog_dems_table <- alldata_melted  %>% 
                  filter(!is.na(cognitive.score), !is.na(dx), !is.na(riskcarrier)) %>%
                  group_by(dx, cog_var, riskcarrier) %>%
                  summarise("Mean" = mean(cognitive.score, na.rm = T),
                            "St Dev" = sd(cognitive.score, na.rm = T))

kable(cog_dems_table)

regress_results <- alldata_melted %>%
                    filter(!is.na(cognitive.score), !is.na(dx), !is.na(riskcarrier)) %>%
                    group_by(cog_var) %>%
                    do(tidy(lm(cognitive.score ~ dx*riskcarrier*age, .)))

kable(regress_results)

#Let's just look at the carier by age interaction term

age_regress_results <- regress_results %>% filter(term == "age") 
age_regress_results$p.fdr = p.adjust(age_regress_results$p.value, method = "fdr")
kable(age_regress_results)

age_regress_results$p.bonferonni = p.adjust(age_regress_results$p.value, method = "bonferroni")

kable(age_regress_results)

```


