---
title: "Stats with R tutorial"
author: "Erin W Dickie"
date: "September 11, 2016"
output: html_document
---

[Return to SCWG Home](https://camh-scwg.github.io/compucool-oct-2016/)

-----

Please download the following files:


1. [messy_demographic.csv](data/messy_demographic.csv) 
2. [messy_cognitive.csv](data/messy_cognitive.csv)
3. [messy_genotype.csv](data/messy_genotype.csv)

----

## Stats sesstion

We need 'rms', 'ggplot2', and 'car' packages
If you haven't already installed these:

Load the packages

```{r}
library(rms)
library(ggplot2)
library(car)
library(dplyr)
```

The data should be merged and ready to go from day 1. If not, here's the code for it:

```{r eval = FALSE}
data1 <- read.csv("~/Downloads/messy_demographic.csv")  # put in the location of the downloaded data file 1
data2 <- read.csv("~/Downloads/messy_cognitive.csv")  # put in the location of the downloaded data file 2
data3 <- read.csv("~/Downloads/messy_genotype.csv")
```
```{r echo = FALSE}
data1 <- read.csv("data/messy_demographic.csv")  # put in the location of the downloaded data file 1
data2 <- read.csv("data/messy_cognitive.csv")  # put in the location of the downloaded data file 2
data3 <- read.csv("data/messy_genotype.csv")
```


```{r}
data1[data1==""] <- NA
data1[data1=="missing"] <- NA
data1[data1=="9999"] <- NA

data1$age <- as.numeric(as.character(data1$age))
data1$ethnicity <- factor(data1$ethnicity,levels=c("Cauc","AA","As","In","Other"))
data1$sex <- factor(data1$sex, levels=c(0,1), labels=c("Male","Female"))
data1$dx <- factor(data1$dx, levels=c(0,1), labels=c("Control","Case"))

data2[data2==""] <- NA
data2[data2=="missing"] <- NA
data2[data2=="9999"] <- NA
data2$cog1 <- as.numeric(as.character(data2$cog1))
data2$cog2 <- as.numeric(as.character(data2$cog2))
data2$cog3 <- as.numeric(as.character(data2$cog3))
data2$subID <- gsub(data2$subID,pattern="subject",replacement="SUB_")

data3[data3==""] <- NA
data3[data3=="missing"] <- NA
data3[data3=="9999"] <- NA
data3$genotype <- factor(data3$genotype, levels=c(0,1,2), labels=c("AA","AG","GG"))
data3$subID <- gsub(data3$subID,pattern="subject",replacement="SUB_")

alldata <- merge(data1,data2,by.x="subject_ID",by.y="subID")
alldata <- merge(alldata,data3,by.x="subject_ID",by.y="subID")
```

# HERE WE GO WITH Statistics!


## RESEARCH AIM 4: Linear Regression
#### total_behaviour_score ~ age 
 
Calculate a composite variable by combining multiple variables
Note: new variables can be made easily (using dplyr's mutate verb)

```{r}
alldata$totalcog <- (alldata$cog1 + alldata$cog3) / alldata$cog2
```


Simple linear regression (two ways: base package and rms)

```{r}
lm.base <- lm(data=alldata, totalcog ~ age)
lm.rms <- ols(data=alldata, totalcog ~ age)
```

Let's compare the output's

```{r}
lm.base
summary(lm.base)
anova(lm.base)
```

**Note**: to make the most out of rms package functionality, we need to store summary stats using the datadist() function. That way, when we call summary() on an ols() object (we just made one called "lm.rms") it will give us useful info. 

```{r}
dd.alldata <- datadist(alldata)
options(datadist="dd.alldata")

lm.rms
summary(lm.rms)
anova(lm.rms)
```

## Visualize results using ggplot2

```{r}
ggplot(data=alldata, aes(y=totalcog, x=age)) + 
	geom_point() + 
	geom_smooth(method=lm)
```

#### Challenge 1: add a title and change the axis labels

Visualize predicted results using rms
```{r}
plot(Predict(lm.rms))
```

Check regression assumption of normal residuals

```{r}
hist(resid(lm.rms))
```

They are not normal! We can look at this formally also:
```{r}
shapiro.test(resid(lm.rms)) 
```
## Let's look at and normalize our outcome...

# two ways of looking at it (base and ggplot2):

```{r}
ggplot(data=alldata, aes(x=totalcog)) + geom_histogram() 
```

Let's try transforming our variable using log and sqrt transformations to see if it helps:

```{r}
alldata$totalcog_log <- log(alldata$totalcog)
alldata$totalcog_sqrt <- sqrt(alldata$totalcog)
```


## Here's a super neat way to optimize power transformations:

The powerTransform function will calculate the best transform for your data.
It saves the best exponent as the value "lambda"

```{r}
# calculate the best exponent using powerTransform:
pT <- powerTransform(alldata$totalcog)
# apply the power transform and save the result to a new variable
alldata$totalcog_pT <- alldata$totalcog^pT$lambda ## note ^ is exponent in r
```

### let's try our regression again with the transformed outcome (using rms):

Note: if we want to use summary() on our ols() object we will have to redefine datadist because we created new variables that were not in the original datadist object

```{r}
dd.alldata <- datadist(alldata)
options(datadist="dd.alldata")
```
#### Challenge 2: run a new regression (using ols) with the your new transformed cognitive score as the dependant variable

#### Challenge 3: use ggplot to make a new plot of this effect

-----------------------

## RESEARCH AIM 5: total_behaviour_score ~ age + sex + genotype (multiple linear regression)

This is were we start to add covariates and do multiple regression

```{r}
lm3 <- ols(data=alldata, totalcog_pT ~ age + genotype + sex)
anova(lm3)
summary(lm3)
```

We can easily visualise the effects of each variable on the outcome using rms, but this feature does not give us plot points and is not very flexible:

```{r}
plot(Predict(lm3))
```

Now let's say we want to recode out genotype variable so that we have only 2 groups: those who "carry" the G allele, and those who do not carry it:

````{r}
alldata$riskcarrier[alldata$genotype=="AG" | alldata$genotype=="GG"] <- "carrier"
alldata$riskcarrier[alldata$genotype=="AA"] <- "non-carrier"
alldata$riskcarrier <- as.factor(alldata$riskcarrier)
```

No we can re-run the model:

```{r}
dd.alldata <- datadist(alldata)
options(datadist="dd.alldata")
lm3 <- ols(data=alldata, totalcog_pT ~ age + sex + riskcarrier)
lm3
anova(lm3)
summary(lm3)
```

## RESEARCH AIM 5: Interaction!
#### total_behaviour_score ~ age*riskcarrier + sex 

The concept of statistical interaction goes by many names and has many definitions.  Simply this is the concept that the effect of one variable changes depending on the value of another variable. 

Interaction is indicated in R formula syntax with a ":" or `*`, depending on if you want to automatically include the main effects of your interacting variables or not. As a general rule, always use `*`.

```{r}
lm4 <- lm(totalcog_pT ~ sex + riskcarrier*age, data=alldata)
summary(lm4)
anova(lm4)
```

---- 


# BONUS SECTION - the "right" way to plot....

## How to visualize a significant effect from our regression

####...Controlling for the other variables in the model....
 
To visualize a given effect more informatively, we want to caculate the residuals of the model lacking our co-varitate of interest and plot those residuals as our outcome:

For genotype we want a boxplot of model residuals:

```{r}
lm3.plot <- ols(data=alldata, totalcog_pT ~ genotype + age)

ggplot(data=alldata, aes(y=resid(lm3.plot), x=sex)) + 
  geom_boxplot()
```

See it thinks NA is a value! That is because the ols() object stores NA input values
as NA residuals, and ggplot2 sees these as another level to plot. Fix by re-running
the model to exclude missing observations and plotting the data subset where NAs are excluded:

```{r}
lm3.plot <- ols(data=subset(alldata,sex!="NA"), totalcog_pT ~ age + genotype)
ggplot(data=subset(alldata,sex!="NA"), aes(y=resid(lm3.plot), x=sex)) + 
  geom_boxplot()
```

Ror age we want a scatterplot of residuals. same subsetting principle applies:

```{r}
lm3.plot <- ols(data=subset(alldata,age!="NA"), totalcog_pT ~ genotype + sex)
ggplot(data=subset(alldata,age!="NA"), aes(y=resid(lm3.plot), x=age)) + 
  geom_point() + 
  geom_smooth(method=lm)
```

**CONCEPT**: notice that if we include age in our model and plot age on the x-axis in our residual plot, the effect is lost - we have modeled it out:

```{r}
lm3.plot <- ols(data=alldata, totalcog_pT ~ genotype + sex + age)
ggplot(data=alldata, aes(y=resid(lm3.plot), x=age)) + 
  geom_point() + 
  geom_smooth(method=lm)
```

To plot the interaction we just tell ggplo2 to plot the regression lines for ou

```{r}
lm4.plot <- ols(data=subset(alldata,age!="NA" & riskcarrier!="NA"), totalcog_pT ~ sex)
ggplot(data=subset(alldata,age!="NA" & riskcarrier!="NA"), aes(y=resid(lm4.plot), x=age, col=riskcarrier)) + 
  geom_point() + 
  geom_smooth(method=lm)
```